Sample-Based Determination of Angular Observables

### 2014-06-24 draft.tex
Slight change of notation in Secs I and II wrt indices

Let $\thvec = (\theta_1, \dots \theta_n)$ denote the set of all
angles, and let $\nuvec = (\nu_1, \dots \nu_m)$ denote the set of all
other non-angular kinematic variables needed to fully specify the
final state of the process under study. For example, $\nuvec$ may
include invariant masses or center-of-mass energies. We define an
angular observable $P_i$ as a coefficient in the probability density
function (PDF) of the process,
\begin{align}
    \label{eq:def-P}
    P(\nuvec, \thvec) = \sum_i P_i(\nuvec) \times f_i(\thvec)\,.
\end{align}
Here, the dependence on the $n$ decay angles $\thvec$ has been
explicitly factored out in terms of the angular functions
$f_i(\thvec)$, which we assume to fulfill the orthonormality relations
\begin{equation}
    \label{eq:def-ortho-rel}
    \int_\Omega f_i(\thvec) f_j(\thvec) \dd^n \theta = \delta_{ij}\,.
\end{equation}
For the purpose of this letter, however, it suffices that the system
of angular functions $f_i(\thvec)$ can be transformed into an
orthonormal basis. The transformation needs to be worked out on a
case-by-case basis. For a selection of $B$ decays, the transformations
are listed in the appendices \ref{app:btokll} through
\ref{app:lambdabtolambdall}.  Note that $P_i$ is \emph{not} a PDF.  \\

For particle decays, $P$ is generally expressed in terms of the fully differential decay width,
\begin{align}
    \label{eq:def-P-decay}
    P(\nuvec, \thvec) \equiv \frac{1}{\Gamma}\frac{\dd^{n+m}\Gamma}{\dd \nu_1 \dots \dd \nu_m\, \dd \theta_1 \dots \theta_n}\,,
\end{align}
where $\Gamma$ is the total decay width. For a scattering process, one can similarly use
\begin{align}
    \label{eq:def-P-scattering}
    P(\nuvec, \thvec) \equiv \frac{1}{\sigma}\frac{\dd^{n+m}\sigma}{\dd \nu_1 \dots \dd \nu_m\, \dd \theta_1 \dots \theta_n}\,,
\end{align}
where the total cross section $\sigma$ is used for the
normalization. Since the determination of the total decay width or
total cross section can be quite difficult, we emphasize that
different normalizations for $P$ can be used.  For instance, the total
decay width (or cross section) of the process of interest can be
replaced by the corresponding quantity of a control-channel
process.\fred{How would $P$ remain a proper probability such that
  $\int P = 1$? You can't simply change the normalization. I propose a
  change of notation: I find it terribly confusing that $P(\nuvec,
  \thvec)$ is a probability but $P_i(\nuvec)$ is not. For $P_i$, later
  on the symbols $J_i, K_i$ etc are used. We don't have to use the
  same symbol everywhere but let's not use $P_i$}\\

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $P_i$
can be projected out of the full PDF $P$ by means of
\begin{equation}
    \label{eq:det-Pi-analytical}
    P_i(\nuvec) = \int_{\thvec = 0}^{2\pi} f_i(\thvec) P(\nuvec, \thvec) \dd^n \theta\,.
\end{equation}
In point of fact, $P_i$ is the $f_i$-th moment of the PDF $P$\fred{This
  notation is uncommon. Is that how we \emph{define} the $f_i$-th
  moment?}. Additionally, integration over the non-angular variables
yields
\begin{equation}
    \langle P_i\rangle
    \equiv \int P_i(\nuvec) \dd^m \nu
    = \int \left[\int_{\thvec=0}^{2\pi} f_i(\thvec) P(\nuvec,\thvec) \dd^n\theta\right]\dd^m \nu\,.
\end{equation}
The remainder of this section  describes the method of moments, in which we replace the
analytical integration \refeq{det-Pi-analytical} by Monte Carlo (MC) estimators.\\

The central tenet of MC integration is the fact that the expectation value $E_P[g]$ of some function
$g(x)$ under the probability density $P(x)$,
\begin{equation}
    E_P[g] \equiv \int g(x) P(x) \dd x
\end{equation}
can be approximated \citeneeded\fred{Robert, Casella} by an MC estimator $\what{E_P[g]}$
\begin{equation}
    \label{eq:mc-id}
    E_P[g] \to \widehat{E_P[g]} \equiv \frac{1}{N} \sum_{k=1} g(x^{(k)})
\end{equation}
due to the strong law of large numbers for $N \to \infty$ assuming
that the variates $x^{(k)}$, $k = 1, \dots, N$ are distributed as
$P$; i.e.,
\begin{equation}
    x^{(k)} \sim P\,.
\end{equation}
Throughout this letter we denote all MC estimators with a wide hat.\\

@@

Application of \refeq{mc-id} then yields \fred{notation is very confusing $P_i$ for $x^{(k)} \sim P$}
\begin{equation}
    \langle P_i\rangle \to \widehat{\langle P_i\rangle} = \frac{1}{N} \sum_{k=1}^{N} f_i(x^{(k)})\,.
\end{equation}
It is often of interest to obtain the $\nuvec$- observables integrated
over certain ranges of $\nuvec$; i.e., binned measurements of $\langle
P_i\rangle$\footnote{% We refer for example to the determination of
  the angular observables $J_i$ in $\bar{B}\to\bar{K}^*\ell^+\ell^-$
  decays, see e.g. \cite{Aaij:2013iag}. For details on the relation
  between the $P_i$ and the angular observable $J_i$ in the
  aforementioned decay, we refer to appendix \ref{app:btokstarll}.  }.
Therefore, we introduce the bin-integrated quantity
\begin{align}
    \langle P_i\rangle_{\vec{a},\vec{b}}
    & = \int_{\vec{a}}^{\vec{b}} P_i(\nuvec) \dd^m \nu\\
    & = \int_{\vec{a}}^{\vec{b}} \left[\int_{\thvec=0}^{2\pi} f_i(\thvec) P(\nuvec,\thvec) \dd^n\theta\right]\dd^m \nu\\
    & = \int \left[\int_{\thvec=0}^{2\pi} f_i(\thvec) P(\nuvec,\thvec)
        \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,
        %% \theta^{(m)}(\nuvec - \vec{a}) \theta^{(m)}(\vec{b} - \nuvec)
        \dd^n\theta\right]\dd^m \nu\,,
\end{align}
where the argument of the indicator function $ \mathbf{1}(\vec{a} \le
\nuvec \le \vec{b})$ is to be interpreted componentwise.
Application of \refeq{mc-id} immediately yields
\begin{equation}
    \label{bin-importance}
    \widehat{\langle P_i\rangle_{\vec{a},\vec{b}}}
    = \frac{1}{N} \sum_{k=1}^{N} f_i(x^{(k)})         \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,.
\end{equation}
This basically reduces the estimation of
$\aver{P_i}_{\vec{a},\vec{b}}$ to a (weighted) counting experiment. In
the limit $N \to \infty$, the random vector $( \dots,
\aver{P_i}_{\vec{a},\vec{b}}, \dots)$ follows a multivariate Gaussian
distribution \citeneeded\fred{Why?  multivariate in $\widehat{P_i}$?}
with a mean vector estimated by \refeq{bin-importance} and covariance
estimated as
\begin{equation}
    \cov[\aver{P_i},\aver{P_j}]_{\vec{a},\vec{b}} \to \what{\cov}[\aver{P}_i, \aver{P}_j]_{\vec{a},\vec{b}}
        = \frac{1}{N - 1} \sum_{k=1}^{N} \big[{f}_i(x^{(k)}) - \widehat{\langle P_i\rangle_{\vec{a},\vec{b}}}\big]\,\big[f_j(x^{(k)}) - \widehat{\langle P_j\rangle_{\vec{a},\vec{b}}}\big]
        \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,.
\end{equation}
We would like to emphasize that the sample covariance should rapidly \fred{So what value of $N$ is close enought to $\infty$?} converge towards the true covariance matrix.
This will allow to construct constraints on optimized observables with correlated uncertainties
from the experimental constraints on the angular observables. (See \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf}
for definitions of such optimized observables in $B\to K^*\ell^+\ell^-$ decays).

>>>>

Let $\thvec = (\theta_1, \dots \theta_N)$ denote the set of all
angles, and let $\nuvec = (\nu_1, \dots \nu_M)$ denote the set of all
other non-angular kinematic variables needed to fully specify the
final state of the process under study. For example, $\nuvec$ may
include invariant masses or center-of-mass energies. We define an
angular observable $P_i$ as a coefficient in the probability density
function (PDF) of the process,
\begin{align}
    \label{eq:def-P}
    P(\nuvec, \thvec) = \sum_i P_i(\nuvec) \times f_i(\thvec)\,.
\end{align}
Here, the dependence on the $n$ decay angles $\thvec$ has been
explicitly factored out in terms of the angular functions
$f_i(\thvec)$, which we assume to fulfill the orthonormality relations
\begin{equation}
    \label{eq:def-ortho-rel}
    \int_\Omega f_i(\thvec) f_j(\thvec) \dd^N \theta = \delta_{ij}\,.
\end{equation}
For the purpose of this letter, however, it suffices that the system
of angular functions $f_i(\thvec)$ can be transformed into an
orthonormal basis. The transformation needs to be worked out on a
case-by-case basis. For a selection of $B$ decays, the transformations
are listed in the appendices \ref{app:btokll} through
\ref{app:lambdabtolambdall}.  Note that $P_i$ is \emph{not} a PDF.  \\

For particle decays, $P$ is generally expressed in terms of the fully differential decay width,
\begin{align}
    \label{eq:def-P-decay}
    P(\nuvec, \thvec) \equiv \frac{1}{\Gamma}\frac{\dd^{N+M}\Gamma}{\dd \nu_1 \dots \dd \nu_M\, \dd \theta_1 \dots \theta_N}\,,
\end{align}
where $\Gamma$ is the total decay width. For a scattering process, one can similarly use
\begin{align}
    \label{eq:def-P-scattering}
    P(\nuvec, \thvec) \equiv \frac{1}{\sigma}\frac{\dd^{N+M}\sigma}{\dd \nu_1 \dots \dd \nu_M\, \dd \theta_1 \dots \theta_N}\,,
\end{align}
where the total cross section $\sigma$ is used for the
normalization. Since the determination of the total decay width or
total cross section can be quite difficult, we emphasize that
different normalizations for $P$ can be used.  For instance, the total
decay width (or cross section) of the process of interest can be
replaced by the corresponding quantity of a control-channel
process.\fred{How would $P$ remain a proper probability such that
  $\int P = 1$? You can't simply change the normalization. I propose a
  change of notation: I find it terribly confusing that $P(\nuvec,
  \thvec)$ is a probability but $P_i(\nuvec)$ is not. For $P_i$, later
  on the symbols $J_i, K_i$ etc are used. We don't have to use the
  same symbol everywhere but let's not use $P_i$}\\

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $P_i$
can be projected out of the full PDF $P$ by means of
\begin{equation}
    \label{eq:det-Pi-analytical}
    P_i(\nuvec) = \int_{\thvec = 0}^{2\pi} f_i(\thvec) P(\nuvec, \thvec) \dd^N \theta\,.
\end{equation}
In point of fact, $P_i$ is the $f_i$-th moment of the PDF $P$\fred{This
  notation is uncommon. Is that how we \emph{define} the $f_i$-th
  moment?}. Additionally, integration over the non-angular variables
yields
\begin{equation}
    \langle P_i\rangle
    \equiv \int P_i(\nuvec) \dd^M \nu
    = \int \left[\int_{\thvec=0}^{2\pi} f_i(\thvec) P(\nuvec,\thvec) \dd^N\theta\right]\dd^M \nu\,.
\end{equation}
The remainder of this section  describes the method of moments, in which we replace the
analytical integration \refeq{det-Pi-analytical} by Monte Carlo (MC) estimators.\\

The central tenet of MC integration is the fact that the expectation value $E_P[g]$ of some function
$g(x)$ under the probability density $P(x)$,
\begin{equation}
    E_P[g] \equiv \int g(x) P(x) \dd x
\end{equation}
can be approximated \citeneeded\fred{Robert, Casella} by an MC estimator $\what{E_P[g]}$
\begin{equation}
    \label{eq:mc-id}
    E_P[g] \to \widehat{E_P[g]} \equiv \frac{1}{K} \sum_{k=1} g(x^{(k)})
\end{equation}
due to the strong law of large numbers for $K \to \infty$ assuming
that the variates $x^{(k)}$, $k = 1, \dots, K$ are distributed as
$P$; i.e.,
\begin{equation}
    x^{(k)} \sim P\,.
\end{equation}
Throughout this letter we denote all MC estimators with a wide hat.\\

@@

Application of \refeq{mc-id} then yields \fred{notation is very confusing $P_i$ for $x^{(k)} \sim P$}
\begin{equation}
    \langle P_i\rangle \to \widehat{\langle P_i\rangle} = \frac{1}{K} \sum_{k=1}^{K} f_i(x^{(k)})\,.
\end{equation}
It is often of interest to obtain the $\nuvec$- observables integrated
over certain ranges of $\nuvec$; i.e., binned measurements of $\langle
P_i\rangle$
Therefore, we introduce the bin-integrated quantity
\begin{align}
    \langle P_i\rangle_{\vec{a},\vec{b}}
    & = \int_{\vec{a}}^{\vec{b}} P_i(\nuvec) \dd^M \nu\\
    & = \int_{\vec{a}}^{\vec{b}} \left[\int_{\thvec=0}^{2\pi} f_i(\thvec) P(\nuvec,\thvec) \dd^N\theta\right]\dd^M \nu\\
    & = \int \left[\int_{\thvec=0}^{2\pi} f_i(\thvec) P(\nuvec,\thvec)
        \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,
        %% \theta^{(m)}(\nuvec - \vec{a}) \theta^{(m)}(\vec{b} - \nuvec)
        \dd^n\theta\right]\dd^m \nu\,,
\end{align}
where the argument of the indicator function $ \mathbf{1}(\vec{a} \le
\nuvec \le \vec{b})$ is to be interpreted componentwise.
Application of \refeq{mc-id} immediately yields
\begin{equation}
    \label{bin-importance}
    \widehat{\langle P_i\rangle_{\vec{a},\vec{b}}}
    = \frac{1}{N} \sum_{k=1}^{K} f_i(x^{(k)})         \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,.
\end{equation}
This basically reduces the estimation of
$\aver{P_i}_{\vec{a},\vec{b}}$ to a (weighted) counting experiment. In
the limit $N \to \infty$, the random vector $( \dots,
\aver{P_i}_{\vec{a},\vec{b}}, \dots)$ follows a multivariate Gaussian
distribution \citeneeded\fred{Why?  multivariate in $\widehat{P_i}$?}
with a mean vector estimated by \refeq{bin-importance} and covariance
estimated as
\begin{equation}
    \cov[\aver{P_i},\aver{P_j}]_{\vec{a},\vec{b}} \to \what{\cov}[\aver{P}_i, \aver{P}_j]_{\vec{a},\vec{b}}
        = \frac{1}{K - 1} \sum_{k=1}^{K} \big[{f}_i(x^{(k)}) - \widehat{\langle P_i\rangle_{\vec{a},\vec{b}}}\big]\,\big[f_j(x^{(k)}) - \widehat{\langle P_j\rangle_{\vec{a},\vec{b}}}\big]
        \mathbf{1}(\vec{a} \le \nuvec \le \vec{b})\,.
\end{equation}
We would like to emphasize that the sample covariance should rapidly \fred{So what value of $K$ is close enought to $\infty$?} converge towards the true covariance matrix.
This will allow to construct constraints on optimized observables with correlated uncertainties
from the experimental constraints on the angular observables. (See \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf}
for definitions of such optimized observables in $B\to K^*\ell^+\ell^-$ decays).
